---
title: "rd4 Basic Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Load dependencies for the vignette

```{r setup}
suppressPackageStartupMessages(library(rd4))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(knitr))
```

## Locate and load the example D4 file

```{r load_d4}
d4_path <- system.file("extdata", "example.d4", package="rd4")
d4source <- D4Source(d4_path)
```

## Explore some basic features of the data

Chromosome names and lengths

```{r chrs}
chrs <- get_chroms(d4source)
head(rbindlist(chrs))
```

Track names

```{r tracks}
get_tracks(d4source)
```

## Explore data values across a genomic region

Region coordinates are 0-based, half-open.

### Summary statistics

```{r summary}
mean(d4source, "chr1", 17027500, 17027600)
median(d4source, chr = "chr1", start = 17027500, end = 17027600)
percentile(d4source, "chr1", 17027500, 17027600, 50)
```

### Vector of data values

```{r query}
# Query a genomic interval
query <- query(d4source, "chr1", 17027500, 17027600)

# Fields in the result object
names(query)

# Vector of data values
query$results

# The query that was supplied
query$query

# The same region on the minus strand
query(d4source, "chr1", 17027500, 17027600, minus_strand =TRUE)$results
```

## Resample a region

This operation reduces the resolution of the data vector, summarizing values at the level of the bin size and returning a new query result.

```{r resample}
query_resample <- resample(
  d4source, "chr1", 17027500, 17027600, method = "mean", bin_size = 10)

# The result has the same structure as a `query()` result
names(query_resample)

# Values are summarized per bin
query_resample$results

# The bin_size field is set
query_resample$bin_size
```

## Working with `GenomicRanges` `Granges` objects

[GenomicRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) is the [Bioconductor](https://bioconductor.org/) package for representation and manipulation of genomic intervals. `rd4` provides methods to batch query a D4 file with all intervals in a `GRanges` object, updating the metadata for each genomic range with the values from the D4 data.

Each method returns a new `GRanges` object with updated metadata.

```{r granges}
# Manually construct a GRanges object with the above genomic interval and one 
# other interval
# `IRanges` coordinates are 1-based inclusive
granges <- GRanges(
  seqnames = "chr1", 
  ranges = IRanges(start = c(17027501, 17027701), end = c(17027600, 17027800)), 
  strand = "+"
)

# Add summary statistics as interval metadata
granges <- update_mean(d4source, granges)
granges <- update_median(d4source, granges)
granges <- update_percentile(d4source, granges, 25)
granges <- update_percentile(d4source, granges, 50)
granges <- update_percentile(d4source, granges, 75)

# View populated metadata
print(granges)
kable(mcols(granges))

# Add position-level values as interval metadata
granges <- update_query_results(d4source, granges)
lapply(mcols(granges)$query_results, function(result) result$results)
```

## Session info

```{r session_info}
sessionInfo()
```
