---
title: "rd4 basic usage"
author: 
  - Pamela Russell
  - Seth Stadick
package: rd4
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{rd4 Basic Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The Dense Depth Data Dump (D4) format provides fast analysis and compact storage of quantitative genomics datasets, and is generally an efficient alternative to existing formats such as bedGraph and BigWig. It supports random access, multiple tracks (e.g. different data types or samples), and summary statistics on arbitrary genomic intervals. For full details on the format, see [Hou et al.](https://doi.org/10.1038/s43588-021-00085-0).

`rd4` is the first package providing R bindings for reading and querying D4 files, and is similar in scope to the existing Python library [pyd4](https://github.com/38/pyd4). `rd4` provides functions to open a D4 file, explore its contents, and extract summary statistics and complete data for arbitrary genomic intervals. 

Additionally, `rd4` enables seamless Bioconductor workflows through support for the standard genomic annotation package [GenomicRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html). D4 files can be queried and exported using functions that take in and return `GRanges` objects.


# Installation

```{r eval = FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("rd4")
```


# Package usage

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Load dependencies for the vignette

```{r setup}
suppressPackageStartupMessages(library(rd4))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(GenomicRanges))
```

## Locate and load the example D4 file

```{r load_d4}
d4_path <- system.file("extdata", "example.d4", package="rd4")
d4source <- D4Source(d4_path)
```

## Explore some basic features of the data

Print object summary

```{r print}
print(d4source)
```

Sequence information

```{r chrs}
seqinfo(d4source)
```

Track names

```{r tracks}
get_tracks(d4source)
```

## Explore data values across a genomic region

Coordinates in the D4 format are 0-based end exclusive, while all coordinates in `rd4` are 1-based, end inclusive.

### Summary statistics

For these and below functions, omitting the track name parameter causes the first track to be queried.

```{r summary_notrack}
mean(d4source, chr = "chr1", start = 17027500, end = 17027600)
median(d4source, chr = "chr1", start = 17027500, end = 17027600)
percentile(d4source, "chr1", 17027500, 17027600, 50)
```

Query a specific track:

```{r summary_track}
mean(d4source, chr = "chr3", start = 0, end = 100000000, track = "track2")
median(d4source, chr = "chr3", start = 0, end = 100000000, track = "track2")
percentile(d4source, "chr3", 0, 100000000, 99, track = "track2")
percentile(d4source, "chr3", 0, 100000000, 99.999, track = "track2")
```

### Vector of data values

```{r query}
# Query a genomic interval
query <- query(d4source, "chr1", 17027500, 17027600)

# Fields in the result object
names(query)

# Vector of data values
query$results

# The query that was supplied
query$query

# The same region on the minus strand
query(d4source, "chr1", 17027500, 17027600, minus_strand =TRUE)$results
```

## Resample a region

This operation reduces the resolution of the data vector, summarizing values at the level of the bin size and returning a new query result.

```{r resample}
query_resample <- resample(
  d4source, "chr1", 17027500, 17027600, method = "mean", bin_size = 10)

# The result has the same structure as a `query()` result
names(query_resample)

# Values are summarized per bin
query_resample$results

# The bin_size field is set
query_resample$bin_size
```

## Working with `GenomicRanges` `Granges` objects

[GenomicRanges](https://bioconductor.org/packages/release/bioc/html/GenomicRanges.html) is the [Bioconductor](https://bioconductor.org/) package for representation and manipulation of genomic intervals. 

### Querying a D4 file with intervals in a `GRanges` object

`rd4` provides methods to batch query a D4 file with all intervals in a `GRanges` object, updating the metadata for each genomic range with the values from the D4 data.

Each method returns a new `GRanges` object with updated metadata.

```{r granges_query}
# Manually construct a GRanges object with the above genomic interval and one 
# other interval
# `IRanges` coordinates are 1-based inclusive
granges <- GRanges(
  seqnames = "chr1", 
  ranges = IRanges(start = c(17027501, 17027701), end = c(17027600, 17027800)), 
  strand = "+"
)

# Add summary statistics as interval metadata
granges <- update_with_mean(granges, d4source)
granges <- update_with_median(granges, d4source)
granges <- update_with_percentile(granges, d4source, 25)
granges <- update_with_percentile(granges, d4source, 50)
granges <- update_with_percentile(granges, d4source, 75)

# View populated metadata
print(granges)
kable(mcols(granges))

# Add position-level values as interval metadata
granges <- update_with_query_results(granges, d4source)
lapply(mcols(granges)$query_results, function(result) result$results)
```

### Converting data profile from a D4 file to a `GRanges` object

The `import()` function converts data in a `D4Source` object to a `GRanges` object. The data values are stored in a metadata column.

The function name and some arguments are analogous to the `import.*()` functions in the [rtracklayer](https://www.bioconductor.org/packages/release/bioc/html/rtracklayer.html) package. As such, specific query region(s) of interest can be supplied as a `GRanges` or `GRangesList` object via the `which` argument, or the whole genome can be queried by omitting `which`.

Note: if data values are continuous or vary a lot, for large regions this function may be slow and may return a very large `GRanges` object. It is recommended to use it only when a `GRanges` object is required to plug into other Bioconductor workflows.

Whole genome:

```{r granges_export_genome}
import(d4source)
```

Specific region:

```{r granges_export_region}
import(
  d4source, 
  which = GRanges(
    seqnames = "chr1", 
    IRanges(start = c(17027501, 17027701), end = c(17027600, 17027800))
  )
)
```

## Session info

```{r session_info}
sessionInfo()
```
